Efficient High-Level Abstractions for Web Programming

### 2013-06-11 papers/gpce2013/gpce2013.tex
gpce draft

\title{How to build an homogeneous language for heterogeneous platforms}
\subtitle{You don’t have to trade abstraction for control}

@@

\begin{abstract}
Writing large Web applications is known to be difficult. One challenge comes from the fact that the application's
logic is scattered into heterogeneous clients and servers, making it difficult to share code between both sides or to
move code from one side to the other. Another challenge is performance: while Web applications rely on ever more
code on the client-side, they may run on smart phones with little hardware capabilities. These two challenges raise
the following problem: how to benefit from high-level languages and libraries making code complexity easier to manage
and abstracting over the clients and servers differences without trading this engineering comfort for performance?
This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
code leveraging the target platform characteristics, (2) be shared between clients and servers. Code written with
our DSL has a high level of abstraction but runs as fast as hand tuned low-level JavaScript code.
\end{abstract}

@@

\keywords Heterogeneous code generation, Domain-specific languages

@@

Web applications are attractive because they require no installation or deployment steps on clients and enable large
scale collaborative experiences. However, writing large Web applications is known to be
difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
that the business logic is scattered into heterogeneous client-side and server-side
environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
engineering process and requires a higher maintenance effort: if a piece of logic is implemented on client-side and
finally needs to be implemented on server-side instead, the code can not be reused and the feature needs to be
completely rewritten (and \emph{vice versa}). Even worse, logic parts that run on both client-side and server-side
need to be duplicated. For instance, HTML fragments may be built from the server-side when a page is requested by a
client, but they may also be built from the client-side to perform an incremental update subsequent to an user
action. How could developers write HTML fragment definitions once and render them on both client-side and
server-side?

@@

Generating efficient code for heterogeneous platforms is hard to achieve in an extensible way: the translation of
common abstractions like collections into their native counterpart (JavaScript arrays on client-side and standard
library's collections on server-side) may be hard coded in the compiler, but that would not scale to handle all the
abstractions a complete application may use (\eg HTML fragment definitions, form validation rules, or even some
business data type that may be represented differently).

@@

Kossakowski \etal introduced \emph{js-scala}, a compiled embedded DSL defined in Scala that generates JavaScript
code, making it possible to write the client-side code of Web applications using
JavaScript~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the engineering dilemma described
above. This paper enriches \emph{js-scala} with the following contributions\footnote{The code is available at
\href{http://github.com/js-scala}{http://github.com/js-scala}}:

@@

We validate our approach with a case study implemented with various candidate technologies and discuss the relative
pro and cons of them. Though the code written in our DSL is high-level and can be shared between clients and servers,
it has the same runtime performances as hand-tuned low-level JavaScript code.

The remainder of this paper is organized as follows. The next section introduces existing approaches for defining
cross-compiling languages. Sections \ref{contribution} and \ref{contribution-shared} present our contribution.
Section \ref{validation} compares our solution to common approaches. Section \ref{discussion} concludes.

@@

The first approach for defining a cross-platform language consists in hard-coding, in the compiler, the
code generation scheme of each language feature to each target platform. Figure \ref{langs} (a) depicts this
process. In order to support a feature related to a specific domain, the whole compiler pipeline (parser, code
generator, \etc) may have to be adapted. This approach gives \emph{fat} languages because a lot of concepts are
defined at the language level: general programming concepts such as naming, functions, classes, and more specific
concepts such as HTML fragment definition. Examples of such languages are Links~\cite{Cooper07_Links}, Opa and
Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because each concept is defined in the compiler,
and modifying a compiler requires a high effort. Furthermore, these languages also require to support common
programming abstraction and composition mechanisms, as general purpose languages do. So they usually try to re-invent
the features of general purpose languages. We argue that this approach for defining programming languages is
difficult to scale: for every problem you have to rewrite a full-featured programming language besides addressing the
concepts specific to the problem domain.

@@

Another approach consists in defining several independent domain-specific languages~\cite{Van00_DSL}, each one
focusing on concerns specific to a given problem domain, and then to combine all the source artifacts written with
these language into one executable program, as shown in figure~\ref{langs} (b). Defining such a language requires a
minimal effort compared to the previous approach. On the other hand, it is difficult to have interoperability between
DSLs (ref). \cite{Groenewegen08_WebDSL} shows such a domain-specific language for defining Web applications.

@@

Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
language (this is also referred to as \emph{embedded} domain-specific languages). Figure \ref{langs} (c) depicts this
approach. The general purpose language is used as a host language and does not need to be modified if a new concept
is introduced, because concepts are defined as pure libraries. However, this approach gives no opportunity to
translate a concept efficiently according to the target platform characteristics. Examples of languages following
this approach are Java/GWT, Kotlin, HaXe and SharpKit.

@@

The last approach, shown in figure \ref{langs} (d), can be seen as a middle-ground between the two previous
approaches: DSLs are embedded in a host language but use a code generation process. The embedding makes it easy to
combine several DSLs and makes the language implementation easier because the host language infrastructure is
reused. On the other hand, the syntax of the DSLs is limited by the syntax flexibility of the host language. The code
generation process is specific to each DSL and gives the opportunity to perform domain-specific optimizations.

@@

LMS is based on staging~\cite{Jorring1986_Staging}: a program using LMS is a regular Scala program that evaluates
to an intermediate representation (IR) of a final program. A code generator can traverse this IR to produce the final
program code. Expressions evaluated in the initial program and those evaluated in the final program are distinguished
by their type: a \code{Rep[Int]} value in the initial program is a staged computation that generates code evaluating
to an \code{Int} value in the final program. An \code{Int} computation in the initial program is evaluated during the
initial program evaluation and becomes a constant in the final program.

@@

Listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} using \emph{js-scala}. We
provide two functions for searching elements: \code{find} to find the first element matching a selector and
\code{findAll} to find all the matching elements. During staging these functions try to analyze the selector that is
passed as parameter and, when appropriate, produce code using the specialized API, otherwise they produce code using
\code{querySelector} and \code{querySelectorAll}. As a result, listing~\ref{js-scala-selectors} generates a
JavaScript program identical to listing~\ref{vanilla-selectors}: the high-level abstraction (the \code{find} and
\code{findAll} functions) exist only in the initial program, not in the final JavaScript program. It gives two
advantages: (1) the execution of the final program is more performant because of the use of specialized APIs, (2) the
final program’s size is smaller because it does not need to include jQuery. (Todo. Move this last sentence in the
discussion)

@@

Listing \ref{selector-impl} shows the implementation of the \code{find} function producing different IR node
according to the selector passed as parameter. The \code{getConstIdCss} function analyzes the selector: if it
is a constant \code{String} value containing a CSS ID selector, it returns the value of the identifier. If the
\code{find} function is applied to the \code{document} and to an ID selector, it returns a
\code{DocumentGetElementById} IR node (that is translated to a \code{document.getElementById} call by the code
generator), otherwise it returns a \code{SelectorFind} IR node (that is translated to a \code{querySelector}
call). (FIXME Add a ref to the spec of CSS selectors?)

@@

Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
example, consider listing \ref{null-unsafe} finding a particular widget in the page and then a particular
button in the widget. The native \code{querySelector} method returns \code{null} if no node matched the given
selector in the document. If we run this code in a page where the widget is not present, it will throw an error
and stop further JavaScript execution. Defensive code can be written to handle \code{null} references, but leads to
very cumbersome code, as shown in listing \ref{null-defensive}.

@@

Listing \ref{null-js-scala} implements in \emph{js-scala} a program equivalent to listing \ref{null-defensive}. The
\code{for} notation is used to sequence computations over optional values encoded with a monad. The \code{find}
function returns a \code{Rep[Option[Element]]} value, that can either be a \code{Rep[Some[Element]]} (if an element
was found) or a \code{Rep[None.type]} (if no element was found). The \code{for} expression contains a sequence
of statements that are executed in order, as long as the previous statement returned a \code{Rep[Some[Element]]}
value.

@@

Such a monadic API brings both safety and expressiveness to developers manipulating optional values but usually
involves the creation of an extra container object holding the optional value. In our case, we expose the monadic API
at the first stage but we generate code that does not wrap values in container objects but instead checks if they are
\code{null} or not when dereferenced. So the extra container object exists only at the stage-level and is removed
during code generation: listing \ref{null-js-scala} produces a code equivalent to listing \ref{null-defensive}.

@@

Last but not least, we leverage the modular architecture of LMS to define two code generators for this API: one
targeting JavaScript and the other targeting Scala. So the same abstraction is efficiently translated on both server
and client sides.

@@

Listings \ref{js-gen-forest} and \ref{scala-gen-forest} show the relevant parts of the code generators for this DSL.
They basically follow the same pattern: they visit \code{Tag} and \code{Text} IR nodes and produce the corresponding
elements in the target language. (TODO More details)

@@

@@

@@

Figure \ref{micro-benchmark} shows the micro-benchmark results. We measured the runtime performance, the number of
lines of code (as an indication of the abstraction level), and finally the ratio of runtime performance and lines of
code. \emph{js-scala} is between 3 to 100 times faster than other approaches on our benchmark. Combined with the
fact that it provides high-level abstractions, it has a performance / lines of code ratio more than 4 times higher
than other approaches.

@@

The application was initially written using jQuery. We rewrote it using several technologies for the client-side
part: vanilla JavaScript (low-level code without third-party library), js-scala, GWT and HaXe. In each case we tried
to write the application in an idiomatic way.

\paragraph*{Vanilla JavaScript} Low-level code.

\paragraph*{jQuery} High-level code.

\paragraph*{js-scala} High-level code. HTML fragment definition reused between server and client sides.

\paragraph*{HaXe} Low-level code.

\paragraph*{GWT} High-level code?

\subsection{Benchmarks, Code Metrics}

Our goal is to evaluate the level of abstraction provided by each solution and their performances. We took the
number of lines of code as a measure of the level of abstraction. We also measured the ability to share code between
client and server sides.

We simulated user actions on a Web page (2000 clicks on buttons, triggering a dynamic update of the page and
involving the use of the Option monad, the selectors API and the HTML fragment definition API) and measured the time
it took complete them. We only measured the execution time of client-side code execution. The tests were run on a
DELL Latitude E6430 laptop with 8 GB of RAM, on the Google Chrome v27 Web browser. Figure \ref{benchmark} shows the
benchmark results.

@@

The runtime performances of the vanilla JavaScript, HaXe and \emph{js-scala} versions are similar (though the
\emph{js-scala} version is slightly slower by 6\%). It is worth noting that the vanilla JavaScript and the HaXe
versions use low-level code compared to \emph{js-scala}, as shown in the second part of the figure (lines of code):
the \emph{js-scala} version needs only 74 lines of code while the vanilla JavaScript version needs 116 lines of code
(57\% bigger) and the HaXe version needs 148 lines of code (100\% bigger). The jQuery JavaScript version, of which
code is high-level (54 lines of code, 27\% less than \emph{js-scala}) runs 10 times slower than the \emph{js-scala}
version.

The last part of the figure compares the runtime performance / lines of code ratio. \emph{js-scala} shows the best
score, being 1.48 times better than the vanilla JavaScript version, 1.88 times better than the HaXe version,
3.77 times better than the GWT version and 7.82 times better than the jQuery JavaScript version.

@@

We implemented a high-level language abstracting over client and server heterogeneity but producing efficient code.

@@

Acknowledgments, if needed.

>>>>

\title{Efficient High-Level Abstractions for Web Programming}
%\subtitle{You don’t have to trade abstraction for control}

@@

\begin{abstract}
Writing large Web applications is known to be difficult. One challenge comes from the fact that the application's
logic is scattered into heterogeneous clients and servers, making it difficult to share code between both sides or to
move code from one side to the other. Another challenge is performance: while Web applications rely on ever more
code on the client-side, they may run on smart phones with little hardware capabilities. These two challenges raise
the following problem: how to benefit from high-level languages and libraries making code complexity easier to manage
and abstracting over the clients and servers differences without trading this engineering comfort for performance?
This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
code leveraging the target platform characteristics, (2) be shared between clients and servers. Though code written
with our DSL has a high level of abstraction our benchmark on a real world application reports that it runs as fast
as hand tuned low-level JavaScript code.
\end{abstract}

@@

\keywords Heterogeneous code generation, Domain-specific languages, Scala, Web

@@

Web applications are attractive because they require no installation or deployment steps on clients and enable large
scale collaborative experiences. However, writing large Web applications is known to be
difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
that the business logic is scattered into heterogeneous client-side and server-side
environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
engineering process and requires a higher maintenance effort: there is no way to move a piece of code targeting
the server-side to target the client-side, the code has to be rewritten. Even worse, logic parts that run on both
client-side and server-side need to be duplicated. For instance, HTML fragments may be built from the server-side
when a page is requested by a client, but they may also be built from the client-side to perform an incremental
update subsequent to an user action. How could developers write HTML fragment definitions once and render them on
both client-side and server-side?

@@

Generating efficient code for heterogeneous platforms is hard to achieve in an extensible way: the translation of
common abstractions like collections into their native counterpart (JavaScript arrays on client-side and standard
library's collections on server-side) may be hard-coded in the compiler, but that approach would not scale to handle
all the abstractions a complete application may use (\eg HTML fragment definitions, form validation rules, or even
some business data type that may be represented differently).

@@

Kossakowski \etal introduced \emph{js-scala}, a compiled embedded DSL defined in Scala that generates JavaScript
code, making it possible to write the client-side code of Web applications using
JavaScript~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the engineering dilemma described
above. This paper enriches js-scala with the following contributions\footnote{The code is available at
\href{http://github.com/js-scala}{http://github.com/js-scala}}:

@@

We validate our approach with a case study implemented with various candidate technologies and discuss the relative
pro and cons of them. Though the code written in our DSL is high-level and can be shared between clients and servers,
it has the same runtime performances on client-side as hand-tuned low-level JavaScript code.

The remainder of this paper is organized as follows. The next section introduces existing approaches for defining
cross-compiling languages and presents the framework we used to define our DSLs. Sections \ref{contribution} and
\ref{contribution-shared} present our contribution. Section \ref{validation} compares our solution to common
approaches. Section \ref{discussion} concludes.

@@

The first approach for defining a cross-platform language consists in hard-coding, in the compiler, the
code generation scheme of each language feature to each target platform. Figure \ref{langs} (a) depicts this
process. In order to support a feature related to a specific domain, the whole compiler pipeline (parser, code
generator, \etc) may have to be adapted. This approach gives \emph{fat} languages because a lot of concepts are
defined at the language level: general programming concepts such as naming, functions, classes, as well as more
domain-specific concepts such as HTML fragment definition. Examples of such languages for Web programming are
Links~\cite{Cooper07_Links}, Opa and Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because
each concept is defined in the compiler, and modifying a compiler requires a high effort. Furthermore, these
languages also require to support common programming abstraction and composition mechanisms, as general purpose
languages do. So they usually try to re-invent the features of general purpose languages, that’s why we argue that
this approach for defining programming languages is difficult to scale: for every specific problem you would have to
rewrite a full-featured programming language besides addressing the concepts specific to the problem domain.

@@

Another approach consists in defining several independent domain-specific languages~\cite{Van00_DSL}, each one
focusing on concerns specific to a given problem domain, and then to combine all the source artifacts written with
these language into one executable program, as shown in figure~\ref{langs} (b). Defining such languages requires a
minimal effort compared to the previous approach because each language has a limited set of features. On the other
hand, it is difficult to have interoperability between DSLs (missing ref). \cite{Groenewegen08_WebDSL} gave an
example of such a domain-specific language for defining Web applications.

@@

Alternatively, one can define concepts relative to a specific domain as a library on top of a thin general purpose
language (it is also referred to as a domain-specific \emph{embedded} language~\cite{Hudak96_DSEL}). Figure
\ref{langs} (c) depicts this approach. The general purpose language is used as a host language and does not need to
be modified if a new concept is introduced, because concepts are defined as pure libraries. On the other hand, the
syntax of the DSLs is limited by the syntax flexibility of the host language. Furthermore, this approach gives no
opportunity to translate a concept efficiently according to the target platform characteristics because concepts are
defined as libraries and are translated by the compiler, which has no domain-specific knowledge. Examples of
languages following this approach are Java/GWT, Kotlin, HaXe and SharpKit.

@@

The last approach, shown in figure \ref{langs} (d), can be seen as a middle-ground between the two previous
approaches: DSLs are embedded in a host language but use a code generation process. This approach share the same
benefits and limits as embedded DSLs for defining language units. The code generation process is specific to each DSL
and gives the opportunity to perform domain-specific optimizations. In other words deeply embedded DSLs bring
domain-specific knowledge to the compiler.

@@

LMS is based on staging~\cite{Jorring1986_Staging}: a program using LMS is a regular Scala program that evaluates
to an intermediate representation (IR) of a final program. This IR is a graph of expressions that can be traversed
by code generators to produce the final program code. Expressions evaluated in the initial program and those
evaluated in the final program (namely, staged expressions) are distinguished by their type: a \code{Rep[Int]} value
in the initial program is a staged expression that generates code evaluating to an \code{Int} value in the final
program. An \code{Int} computation in the initial program is evaluated during the initial program evaluation and
becomes a constant in the final program.

@@

Listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} using js-scala. We
provide two functions for searching elements: \code{find} to find the first element matching a selector and
\code{findAll} to find all the matching elements. During the first evaluation step, these functions try to analyze
the selector that is passed as parameter and, when appropriate, produce code using the specialized  API, otherwise
they produce  code using \code{querySelector} and \code{querySelectorAll}. As a result,
listing~\ref{js-scala-selectors} generates a JavaScript program identical to listing~\ref{vanilla-selectors}: the
high-level abstraction (the \code{find} and \code{findAll} functions) exist only in the initial program, not in the
final JavaScript program.

@@

Listing \ref{selector-impl} shows the implementation of the \code{find} function producing different IR nodes
according to the selector passed as parameter. The \code{getConstIdCss} function analyzes the selector: if it
is a constant \code{String} value containing a CSS ID selector, it returns the value of the identifier. If the
\code{find} function is applied to the \code{document} and to an ID selector, it returns a
\code{DocumentGetElementById} IR node (that is translated to a \code{document.getElementById} call by the code
generator), otherwise it returns a \code{SelectorFind} IR node (that is translated to a \code{querySelector}
call). (FIXME Add a ref to the spec of CSS selectors?)

@@

Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
example, consider listing \ref{null-unsafe} finding a particular widget in the page and then a particular
button within the widget. The native \code{querySelector} method returns \code{null} if no node matched the given
selector in the document. If we run this code in a page where the widget is not present, it will throw an error
and stop further JavaScript execution. Defensive code can be written to handle \code{null} references, but leads to
very cumbersome code, as shown in listing \ref{null-defensive}.

@@

Listing \ref{null-js-scala} implements in js-scala a program equivalent to listing \ref{null-defensive}. The
\code{for} notation is used to sequence computations over optional values that are encoded with a monad. The
\code{find} function returns a \code{Rep[Option[Element]]} value, that can either be a \code{Rep[Some[Element]]} (if
an element was found) or a \code{Rep[None.type]} (if no element was found). The \code{for} expression contains a
sequence of statements that are executed in order, as long as the previous statement returned a
\code{Rep[Some[Element]]} value.

@@

Such a monadic API brings both safety and expressiveness to developers manipulating optional values but usually
involves the creation of an extra container object holding the optional value. In our case, the monadic API is
used in the initial program but generates code that does not wrap values in container objects but instead checks if
they are \code{null} or not when dereferenced. So the extra container object exists only in the initial program and
is removed during code generation: listing \ref{null-js-scala} produces a code equivalent to listing
\ref{null-defensive}.

@@

Last but not least, we make this abstraction available on server-side by writing a code generator similar to the
JavaScript code generator, but targeting Scala (FIXME Show the code generator?). So the same abstraction is
efficiently translated on both server and client sides.

@@

Listings \ref{js-gen-forest} and \ref{scala-gen-forest} show the relevant parts of the code generators for this DSL.
They basically follow the same pattern: they visit \code{Tag} and \code{Text} IR nodes and produce the corresponding
elements in the target language. (FIXME More details)

@@

Our goal was to evaluate the level of abstraction provided by each solution and their performances. We took the
number of lines of code as a measure of the level of abstraction. We also measured the ability to share code between
client and server sides.

The tests were run on a DELL Latitude E6430 laptop with 8 GB of RAM, on the Google Chrome v27 Web browser.

@@

We reimplemented the optional value abstraction and a same program using it in the following languages: JavaScript,
HaXe and Java.

@@

Figure \ref{micro-benchmark} shows the micro-benchmark results. Js-scala is between 3 to 100 times faster than
other approaches on our benchmark. Combined with the fact that it provides high-level abstractions, it has a
performance / lines of code ratio more than 4 times higher than other approaches.

@@

The application was initially written using jQuery. We rewrote it using several technologies for the client-side
part: vanilla JavaScript (low-level code without third-party library), js-scala, GWT and HaXe. In each case we
tried to write the application in an idiomatic way\footnote{Source code is available at
\href{http://github.com/julienrf/chooze}{http://github.com/julienrf/chooze}}.

\subsubsection{Performance}

The benchmark code simulates user actions on a Web page (2000 clicks on buttons, triggering a dynamic update of
the page and involving the use of the optional value monad, the selectors API and the HTML fragment definition API).
Figure \ref{benchmark} shows the benchmark results.

@@

The runtime performances of the vanilla JavaScript, HaXe and js-scala versions are similar (though the js-scala
version is slightly slower by 6\%). It is worth noting that the vanilla JavaScript and the HaXe versions use
low-level code compared to js-scala, as shown in the second part of the figure (lines of code): the js-scala version
needs only 74 lines of code while the vanilla JavaScript version needs 116 lines of code (57\% bigger) and the HaXe
version needs 148 lines of code (100\% bigger). The jQuery JavaScript version, of which code is high-level (54 lines
of code, 27\% less than js-scala) runs 10 times slower than the js-scala version.

The last part of the figure compares the runtime performance / lines of code ratio. Js-scala shows the best score,
being 1.48 times better than the vanilla JavaScript version, 1.88 times better than the HaXe version, 3.45 times
better than the GWT version and 7.82 times better than the jQuery JavaScript version.

\subsubsection{Code Reuse}

We were able to share HTML DOM fragment definitions between server-side and client-side in js-scala.

@@

We implemented a high-level language abstracting over client and server heterogeneity but producing efficient code.
Generated code size?

@@

This work was funded by Zenexity.
